Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2022-04-24 analysis/paper/paper.Rmd
fix reference keys

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
proteins [@hendyProteomicCalculus2018; @warinnerDirectEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculusCooking2014]. 
This dietary information can be preserved within the mineralised dental plaque 
over many millennia, providing a unique window into the food-related behaviours of 
past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncientCalcified2013; @warinnerPathogensHost2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsightsConsumption2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
and nicotine use [@eerkensDentalCalculusSource2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015], 
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
dental calculus to reconstruct diet is a highly unpredictable process.

@@

Statistical analysis was conducted in `r version$version.string`  [@R-base] and
the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
here [@R-here], and patchwork [@R-patchwork].
<!-- R packages used -->

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot)A);
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

[@henryNeanderthalCalculus2014; @buckleyDentalCalculusCooking2014; @wuDietEarliest2021]. 

@@

[@radiniFoodMultiplePathways2017; @radiniMedievalWomenEarly2019].  

>>>>

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
proteins [@hendyProteomicCalculus2018; @warinnerEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculus2014]. 
This dietary information can be preserved within the mineralised dental plaque 
over many millennia, providing a unique window into the food-related behaviours of 
past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentification1975], 
and later by Middleton and Rovner [-@middletonOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncient2013; @warinnerPathogensHost2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsights2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculus2014; @hendyProteomicCalculus2018], 
and nicotine use [@eerkensDentalCalculus2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodPathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015], 
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
dental calculus to reconstruct diet is a highly unpredictable process.

@@

Statistical analysis was conducted in `r version$version.string`  [@Rbase] and
the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
here [@R-here], and patchwork [@R-patchwork].
<!-- R packages used -->

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot)A);
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeography2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

[@henryNeanderthalCalculus2014; @buckleyDentalCalculus2014; @wuDietEarliest2021]. 

@@

[@radiniFoodPathways2017; @radiniMedievalWomenEarly2019].  

### 2022-04-24 analysis/paper/paper.tex
fix reference keys

Statistical analysis was conducted in R version 4.1.2 (2021-11-01) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-adlerSequencingAncientCalcified2013}{}}%
Adler, C. J., Dobney, K., Weyrich, L. S., Kaidonis, J., Walker, A. W., Haak, W., Bradshaw, C. J., Townsend, G., Soltysiak, A., Alt, K. W., et al. (2013). Sequencing ancient calcified dental plaque shows changes in oral microbiota with dietary shifts of the {Neolithic} and {Industrial} revolutions. \emph{Nature Genetics} 45, 450--5, 455e1. doi:\href{https://doi.org/10.1038/ng.2536}{10.1038/ng.2536}.

\leavevmode\vadjust pre{\hypertarget{ref-armitageExtractionIdentificationOpal1975}{}}%
Armitage, P. L. (1975). The {Extraction} and {Identification} of {Opal Phytoliths} from the {Teeth} of {Ungulates}. \emph{Journal of Archaeological Science} 2, 187--197.

@@

\leavevmode\vadjust pre{\hypertarget{ref-buckleyDentalCalculusCooking2014}{}}%
Buckley, S., Usai, D., Jakob, T., Radini, A., and Hardy, K. (2014). Dental {Calculus Reveals Unique Insights} into {Food Items}, {Cooking} and {Plant Processing} in {Prehistoric Central Sudan}. \emph{PLOS ONE} 9, e100808. doi:\href{https://doi.org/10.1371/journal.pone.0100808}{10.1371/journal.pone.0100808}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-dudgeonDietGeographyDrinking2014}{}}%
Dudgeon, J. V., and Tromp, M. (2014). Diet, {Geography} and {Drinking Water} in {Polynesia}: {Microfossil Research} from {Archaeological Human Dental Calculus}, {Rapa Nui} ({Easter Island}). \emph{International Journal of Osteoarchaeology} 24, 634--648. doi:\href{https://doi.org/10.1002/oa.2249}{10.1002/oa.2249}.

\leavevmode\vadjust pre{\hypertarget{ref-eerkensDentalCalculusSource2018}{}}%
Eerkens, J. W., Tushingham, S., Brownstein, K. J., Garibay, R., Perez, K., Murga, E., Kaijankoski, P., Rosenthal, J. S., and Gang, D. R. (2018). Dental calculus as a source of ancient alkaloids: {Detection} of nicotine by {LC-MS} in calculus samples from the {Americas}. \emph{Journal of Archaeological Science: Reports} 18, 509--515. doi:\href{https://doi.org/10.1016/j.jasrep.2018.02.004}{10.1016/j.jasrep.2018.02.004}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-mickleburghNewInsightsConsumption2012}{}}%
Mickleburgh, H. L., and Pagán-Jiménez, J. R. (2012). New insights into the consumption of maize and other food plants in the pre-{Columbian Caribbean} from starch grains trapped in human dental calculus. \emph{Journal of Archaeological Science} 39, 2468--2478. doi:\href{https://doi.org/10.1016/j.jas.2012.02.020}{10.1016/j.jas.2012.02.020}.

\leavevmode\vadjust pre{\hypertarget{ref-middletonExtractionOpalPhytoliths1994}{}}%
Middleton, W. D., and Rovner, I. (1994). Extraction of {Opal Phytoliths} from {Herbivore Dental Calculus}. \emph{Journal of Archaeological Science} 21, 469--473. doi:\href{https://doi.org/10.1006/jasc.1994.1046}{10.1006/jasc.1994.1046}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-R-base}{}}%
R Core Team (2020). \emph{R: {A} language and environment for statistical computing}. {Vienna, Austria}: {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}.

\leavevmode\vadjust pre{\hypertarget{ref-radiniFoodMultiplePathways2017}{}}%
Radini, A., Nikita, E., Buckley, S., Copeland, L., and Hardy, K. (2017). Beyond food: {The} multiple pathways for inclusion of materials into ancient dental calculus. \emph{American Journal of Physical Anthropology} 162, 71--83. doi:\href{https://doi.org/10.1002/ajpa.23147}{10.1002/ajpa.23147}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-warinnerDirectEvidenceMilk2014}{}}%
Warinner, C., Hendy, J., Speller, C., Cappellini, E., Fischer, R., Trachsel, C., Arneborg, J., Lynnerup, N., Craig, O. E., Swallow, D. M., et al. (2014a). Direct evidence of milk consumption from ancient human dental calculus. \emph{Scientific Reports} 4, 7104. doi:\href{https://doi.org/10.1038/srep07104}{10.1038/srep07104}.

@@

{\centering \includegraphics[width=0.7\linewidth]{/mnt/hogwarts/NextCloud/Uni/publications/PhD/byocstarch/analysis/figures/protocol_overview}

@@

{\centering \includegraphics[width=0.5\linewidth]{/mnt/hogwarts/NextCloud/Uni/publications/PhD/byocstarch/analysis/figures/starches_w_bar}

>>>>

Statistical analysis was conducted in R version 4.1.3 (2022-03-10) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-adlerSequencingAncient2013}{}}%
Adler, C. J., Dobney, K., Weyrich, L. S., Kaidonis, J., Walker, A. W., Haak, W., Bradshaw, C. J., Townsend, G., Soltysiak, A., Alt, K. W., et al. (2013). Sequencing ancient calcified dental plaque shows changes in oral microbiota with dietary shifts of the {Neolithic} and {Industrial} revolutions. \emph{Nature Genetics} 45, 450--5, 455e1. doi:\href{https://doi.org/10.1038/ng.2536}{10.1038/ng.2536}.

\leavevmode\vadjust pre{\hypertarget{ref-armitageExtractionIdentification1975}{}}%
Armitage, P. L. (1975). The {Extraction} and {Identification} of {Opal Phytoliths} from the {Teeth} of {Ungulates}. \emph{Journal of Archaeological Science} 2, 187--197.

@@

\leavevmode\vadjust pre{\hypertarget{ref-buckleyDentalCalculus2014}{}}%
Buckley, S., Usai, D., Jakob, T., Radini, A., and Hardy, K. (2014). Dental {Calculus Reveals Unique Insights} into {Food Items}, {Cooking} and {Plant Processing} in {Prehistoric Central Sudan}. \emph{PLOS ONE} 9, e100808. doi:\href{https://doi.org/10.1371/journal.pone.0100808}{10.1371/journal.pone.0100808}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-dudgeonDietGeography2014}{}}%
Dudgeon, J. V., and Tromp, M. (2014). Diet, {Geography} and {Drinking Water} in {Polynesia}: {Microfossil Research} from {Archaeological Human Dental Calculus}, {Rapa Nui} ({Easter Island}). \emph{International Journal of Osteoarchaeology} 24, 634--648. doi:\href{https://doi.org/10.1002/oa.2249}{10.1002/oa.2249}.

\leavevmode\vadjust pre{\hypertarget{ref-eerkensDentalCalculus2018}{}}%
Eerkens, J. W., Tushingham, S., Brownstein, K. J., Garibay, R., Perez, K., Murga, E., Kaijankoski, P., Rosenthal, J. S., and Gang, D. R. (2018). Dental calculus as a source of ancient alkaloids: {Detection} of nicotine by {LC-MS} in calculus samples from the {Americas}. \emph{Journal of Archaeological Science: Reports} 18, 509--515. doi:\href{https://doi.org/10.1016/j.jasrep.2018.02.004}{10.1016/j.jasrep.2018.02.004}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-mickleburghNewInsights2012}{}}%
Mickleburgh, H. L., and Pagán-Jiménez, J. R. (2012). New insights into the consumption of maize and other food plants in the pre-{Columbian Caribbean} from starch grains trapped in human dental calculus. \emph{Journal of Archaeological Science} 39, 2468--2478. doi:\href{https://doi.org/10.1016/j.jas.2012.02.020}{10.1016/j.jas.2012.02.020}.

\leavevmode\vadjust pre{\hypertarget{ref-middletonOpalPhytoliths1994}{}}%
Middleton, W. D., and Rovner, I. (1994). Extraction of {Opal Phytoliths} from {Herbivore Dental Calculus}. \emph{Journal of Archaeological Science} 21, 469--473. doi:\href{https://doi.org/10.1006/jasc.1994.1046}{10.1006/jasc.1994.1046}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Rbase}{}}%
R Core Team (2020). \emph{R: {A} language and environment for statistical computing}. {Vienna, Austria}: {R Foundation for Statistical Computing}; {R Foundation for Statistical Computing}.

\leavevmode\vadjust pre{\hypertarget{ref-radiniFoodPathways2017}{}}%
Radini, A., Nikita, E., Buckley, S., Copeland, L., and Hardy, K. (2017). Beyond food: {The} multiple pathways for inclusion of materials into ancient dental calculus. \emph{American Journal of Physical Anthropology} 162, 71--83. doi:\href{https://doi.org/10.1002/ajpa.23147}{10.1002/ajpa.23147}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-warinnerEvidenceMilk2014}{}}%
Warinner, C., Hendy, J., Speller, C., Cappellini, E., Fischer, R., Trachsel, C., Arneborg, J., Lynnerup, N., Craig, O. E., Swallow, D. M., et al. (2014a). Direct evidence of milk consumption from ancient human dental calculus. \emph{Scientific Reports} 4, 7104. doi:\href{https://doi.org/10.1038/srep07104}{10.1038/srep07104}.

@@

{\centering \includegraphics[width=0.7\linewidth]{/mnt/hogwarts/Insync/publications/PhD/byocstarch/analysis/figures/protocol_overview}

@@

{\centering \includegraphics[width=0.5\linewidth]{/mnt/hogwarts/Insync/publications/PhD/byocstarch/analysis/figures/starches_w_bar}
